---
title: "Appendix"
output: pdf_document
---

\setcounter{page}{22}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(magrittr)
library(fpp3)
library(caret)
library(corrplot)
```


```{r loading data}
mta2020 <- read.csv("https://raw.githubusercontent.com/okhaimova/DATA698/main/MTA_Daily_Ridership_Data__Beginning_2020.csv")
gas <- read.csv("https://raw.githubusercontent.com/okhaimova/DATA698/main/NYC%20Weekly%20Retail%20Gasoline%20and%20Diesel.csv",
                skip = 2, header = TRUE)
bicycle_count <- read.csv("Bicycle_Counts.csv")
counters <- read.csv("Bicycle_Counters.csv")
```


```{r mta cleanup}
mta2020 <- mta2020 %>%
  set_colnames(c("Date", "Subway", "Subway_Pre", "Buses", "Buses_Pre", "LIRR", "LIRR_Pre", 
                 "MetroNorth","MetroNorth_Pre", "Access-A-Ride", "Access-A-Ride_Pre", 
                 "Bridges_and_Tunnels", "Bridges_and_Tunnels_Pre")) %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y")) %>%
  arrange(Date) 
```



```{r gas cleanup}
gas <- gas %>%
  dplyr::select(1:9) %>%
  select(contains(c("Date","City"))) %>%
  set_colnames(c("Date", "All_Grades", "Regular", "Midgrade", "Premium")) %>%
  mutate(Date = as.Date(Date, format = "%b %d, %Y")) %>%
  na.omit()
```


```{r bike cleanup, message = FALSE}
# these are the ids that match the five bridge counters
bridge_ids <- c("100010018", "100062893", "100009428", "300020904", "100009427")

counters <- counters %>%
  mutate(id = as.character(id))

bicycle_count <- bicycle_count %>%
  mutate(date = as.Date(date, "%m/%d/%Y")) %>%
  group_by(id, date) %>%
  summarise(total = sum(counts)) %>%
  filter(id %in% bridge_ids) %>%
  mutate(id = as.character(id)) %>%
  left_join(counters, by = 'id') %>%
  select(c(date, id, name, total)) %>%
  filter(date > "2014-12-31")

brooklyn_febraury_avg <- bicycle_count %>%
  filter(id == "300020904",
         date >= "2022-02-01" & date < "2022-03-01",
         date != "2022-02-08") %>%
  summarise(mean = mean(total)) %>%
  select(mean) %>%
  as.numeric()

# replace the outlier
bicycle_count <- bicycle_count %>%
  mutate(total = replace(total, total == 68200, round(brooklyn_febraury_avg,0)))
```


```{r weekly combine, warning=FALSE, message=FALSE}
mta_weekly <- mta2020 %>%
  filter(Date > "2020-03-01") %>%
  mutate(Date = yearweek(Date)) %>%
  group_by(Date) %>%
  summarise(Subway = sum(Subway),
            Buses = sum(Buses),
            LIRR = sum(LIRR),
            MetroNorth = sum(MetroNorth),
            `Access-A-Ride` = sum(`Access-A-Ride`),
            Bridges_and_Tunnels = sum(Bridges_and_Tunnels)) %>%
  mutate(MTA = rowSums(across(c(Subway, Buses))))
  
gas_weekly <- gas %>%
  filter(Date > "2020-03-01") %>%
  mutate(Date = yearweek(Date))

bike_weekly <- bicycle_count %>%
  filter(date > "2020-03-01") %>%
  mutate(date = yearweek(date)) %>%
  group_by(date) %>%
  summarise(Bikes = sum(total)) %>%
  rename(Date = date) 


weekly <- plyr::join_all(list(mta_weekly, bike_weekly, gas_weekly), 
              by = "Date", type = 'left')

```

#### A.1 

Weekly Price of Gasoline

```{r, out.width = "50%", fig.show="hold"}
gas %>%
  filter(Date >= "2020-01-01") %>%
  ggplot(aes(x = Date, y = All_Grades)) +
  geom_line() +
  labs(title = "Price of Gasoline for All Grades, NYC",
       y = "$USD")

gas %>%
  filter(Date >= "2020-01-01") %>%
  pivot_longer(-Date) %>%
  filter(name != "All_Grades") %>%
  ggplot(aes(x = Date, y = value, colour = name)) +
  geom_line() +
  facet_grid(factor(name, levels =c("Regular", "Midgrade", "Premium")) ~., scales = "free_y") +
  labs(title = "Price of Gasoline, NYC, bu subgrade",
       y = "$USD",
       color = "Gasoline Grade")
```


#### A.2

Daily MTA Ridership

```{r daily mta, message=FALSE, warning=FALSE, fig.height=4, fig.align = 'center'}
mta2020 %>%
  dplyr::select(Date, Subway, Buses) %>%
  pivot_longer(-Date) %>%
  ggplot(aes(x = Date, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~., scales = "free_y") +
  labs(title = "Daily MTA Ridership",
       y = "Count",
       color = "Mode")

mta2020 %>%
  dplyr::select(Date, LIRR, MetroNorth, "Access-A-Ride") %>%
  pivot_longer(-Date) %>%
  ggplot(aes(x = Date, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~., scales = "free_y") +
  labs(title = "Daily MTA Ridership",
       y = "Count",
       color = "Mode")
```

#### A.3 

Weekly MTA Ridership

```{r weekly mta, message=FALSE, warning = FALSE, fig.height=4, fig.align = 'center'}
weekly %>%
  dplyr::select(Date, Subway, Buses, LIRR, MetroNorth, "Access-A-Ride") %>%
  pivot_longer(-Date) %>%
  ggplot(aes(x = Date, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~., scales = "free_y") +
  labs(title = "Weekly MTA Ridership",
       y = "Count",
       color = "Mode")
```

\newpage

#### A.4

STL Decomposition of Subway Ridership

```{r subway stl, message=FALSE, warning = FALSE, fig.height=4, fig.align = 'center'}
mta2020 %>%
  as_tsibble(index = Date) %>%
  model(STL(Subway ~ season(window = "periodic"), robust = TRUE)) %>%
  components() %>%
  autoplot() +
  labs(title = "STL Decomposition of Subway Ridership")
```


#### A.5 

Daily Bicycle Ridership by NYC Bridge

```{r bike, fig.height=4, fig.align = 'center'}
bridge.labs<- c('Brooklyn', 'Queensboro', 'Manhattan', 'Pulaski', 'Williamsburg')

bicycle_count %>%
  mutate(name = replace(name, name == 'Williamsburg Bridge Bike Path', 'Williamsburg'),
         name = replace(name, name == 'Pulaski Bridge', 'Pulaski'),
         name = replace(name, name == 'Manhattan Bridge Bike Comprehensive', 'Manhattan'),
         name = replace(name, name == 'Ed Koch Queensboro Bridge Shared Path', 'Queensboro'),
         name = replace(name, name == 'Comprehensive Brooklyn Bridge Counter', 'Brooklyn')) %>%
  ggplot(aes(x = date, y = total, color = name)) +
  geom_line() +
  facet_grid(name ~., scales = "free_y") +
  ggtitle("Daily Bicycle Ridership by NYC Bridge")
```

#### A.6 

STL Decomposition of Weekly Bicycle Ridership

```{r bike stl, fig.height=4, fig.align = 'center'}
weekly %>%
  as_tsibble(index = Date) %>%
  model(STL(Bikes ~ season(window = "periodic"), robust = TRUE)) %>%
  components() %>%
  autoplot() +
  labs(title = "Bicycle STL Decomposition")
```

\newpage

#### A.7

Gas vs. Mode of Transportation

```{r gas vs, message = FALSE, warning = FALSE, fig.show='hold', out.width="50%"}
weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = MTA), color = "steelblue") +
  geom_line(aes(y = All_Grades * 10000000), color = "darkred") +
  scale_y_continuous(name = "MTA Subway & Bus Ridership",
                     sec.axis = sec_axis(~./10000000, name = "Price of Gasoline")) +
  labs(title = "Gasoline vs. MTA Bus & Subway")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = Buses), color = "steelblue") +
  geom_line(aes(y = All_Grades * 5000000), color = "darkred") +
  scale_y_continuous(name = "MTA Bus Ridership",
                     sec.axis = sec_axis(~./5000000, name = "Price of Gasoline")) +
  labs(title = "Gasoline vs Bus Ridership")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = Subway), color = "steelblue") +
  geom_line(aes(y = All_Grades * 5000000), color = "darkred") +
  scale_y_continuous(name = "MTA Subway Ridership",
                     sec.axis = sec_axis(~./5000000, name = "Price of Gasoline")) +
  labs(title = "Gasoline vs Subway Ridership")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = Bikes), color = "steelblue") +
  geom_line(aes(y = All_Grades * 100000), color = "darkred") +
  scale_y_continuous(name = "Bike Ridership",
                     sec.axis = sec_axis(~./100000, name = "Price of Gasoline")) +
  labs(title = "Gasoline vs Bike Ridership")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = LIRR), color = "steelblue") +
  geom_line(aes(y = All_Grades * 500000), color = "darkred") +
  scale_y_continuous(name = "LIRR Ridership",
                     sec.axis = sec_axis(~./500000, name = "Price of Gasoline")) +
  labs(title = "Gasoline vs LIRR Ridership")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = MetroNorth), color = "steelblue") +
  geom_line(aes(y = All_Grades * 500000), color = "darkred") +
  scale_y_continuous(name = "MetroNorth Ridership",
                     sec.axis = sec_axis(~./500000, name = "Price of Gasoline")) +
  labs(title = "Gasoline vs Metro-North Ridership")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = Bridges_and_Tunnels), color = "steelblue") +
  geom_line(aes(y = All_Grades * 1000000), color = "darkred") +
  scale_y_continuous(name = "Crossings",
                     sec.axis = sec_axis(~./1000000, name = "Price of Gasoline")) +
  labs(title = "Gasoline vs Bridges & Tunnels Crossings")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = `Access-A-Ride`), color = "steelblue") +
  geom_line(aes(y = All_Grades * 50000), color = "darkred") +
  scale_y_continuous(name = "ccess-A-Ride",
                     sec.axis = sec_axis(~./50000, name = "Price of Gasoline")) +
  labs(title = "Gasoline vs Access-A-Ride")
```

\newpage

#### A.8

Baseline Regression Models with Mode of Transportation

###### Subway Linear

```{r}
model1<- lm(Subway ~ All_Grades, weekly)

summary(model1)

```

##### Buses Linear

```{r}
lm(Buses ~ All_Grades, weekly) %>% summary()
```

\newpage

##### LIRR Linear

```{r}
lm(LIRR ~ All_Grades, weekly) %>% summary()
```

##### Metro-North Linear

```{r}
lm(MetroNorth ~ All_Grades, weekly) %>% summary()
```

\newpage

##### Access-a-Ride Linear

```{r}
lm(`Access-A-Ride` ~ All_Grades, weekly) %>% summary()
```

##### Bridges and Tunnels Linear

```{r}
lm(Bridges_and_Tunnels ~ All_Grades, weekly) %>% summary()
```

\newpage

##### Subway and Buses Combined Linear

```{r}
lm(MTA ~ All_Grades, weekly) %>% summary()
```

##### Bikes Linear

```{r}
lm(Bikes ~ All_Grades, weekly) %>% summary()
```

\newpage

#### A.9 

Linear Regression Comparison

```{r}
weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ All_Grades, data = weekly)) %>%
  map(glance) %>%
  do.call(rbind.data.frame, .) %>%
  rownames_to_column() 



weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ All_Grades, data = weekly)) %>% 
  map(glance) %>% 
  do.call(rbind.data.frame, .) %>% 
  rownames_to_column %>% 
  as_tibble %>% 
  ggplot(aes(x = reorder(rowname, -r.squared), y = r.squared)) +
  geom_col() +
  coord_flip() + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Mode of Transportation",
       y = expression( R^{2}),
       title = expression(paste("Comparison of ", R^{2})))
```

\newpage

#### A.10

Baseline Regression Models, Breakdown by Gasoline Grade

###### Subway Linear, by subgrade

```{r}
model1<- lm(Subway ~ Regular + Midgrade + Premium, weekly)

summary(model1)

```

##### Buses Linear, by subgrade

```{r}
lm(Buses ~ Regular + Midgrade + Premium, weekly) %>% summary()
```

\newpage

##### LIRR Linear, by subgrade

```{r}
lm(LIRR ~ Regular + Midgrade + Premium, weekly) %>% summary()
```

##### Metro-North Linear, by subgrade

```{r}
lm(MetroNorth ~ Regular + Midgrade + Premium, weekly) %>% summary()
```

\newpage

##### Access-a-Ride Linear, by subgrade

```{r}
lm(`Access-A-Ride` ~ Regular + Midgrade + Premium, weekly) %>% summary()
```

##### Bridges and Tunnels Linear, by subgrade

```{r}
lm(Bridges_and_Tunnels ~ Regular + Midgrade + Premium, weekly) %>% summary()
```

\newpage

##### Subway and Buses Combined Linear, by subgrade

```{r}
lm(MTA ~ Regular + Midgrade + Premium, weekly) %>% summary()
```

##### Bikes Linear, by subgrade

```{r}
lm(Bikes ~ Regular + Midgrade + Premium, weekly) %>% summary()
```

\newpage

#### A.11

Linear Regression Comparison, by Gasoline Subgrade

```{r}
weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ Regular + Midgrade + Premium, data = weekly)) %>%
  map(glance) %>%
  do.call(rbind.data.frame, .) %>%
  rownames_to_column() 



weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ Regular + Midgrade + Premium, data = weekly)) %>% 
  map(glance) %>% 
  do.call(rbind.data.frame, .) %>% 
  rownames_to_column %>% 
  as_tibble %>% 
  ggplot(aes(x = reorder(rowname, -adj.r.squared), y = adj.r.squared)) +
  geom_col() +
  coord_flip() + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Mode of Transportation",
       y = expression(paste("Adjusted ", R^{2})),
       title = expression(paste("Comparison of Adjusted ", R^{2})))
```

\newpage

#### A.12

Linear-Log Regression with Logarithmic Transformation of Gas

###### Subway Linear-Log

```{r}
lm(Subway ~ log(All_Grades), weekly) %>% summary()

```

##### Buses Linear-Log

```{r}
lm(Buses ~ log(All_Grades), weekly) %>% summary()
```

\newpage

##### LIRR Linear-Log

```{r}
lm(LIRR ~ log(All_Grades), weekly) %>% summary()
```

##### Metro-North Linear-Log

```{r}
lm(MetroNorth ~ log(All_Grades), weekly) %>% summary()
```

\newpage

##### Access-a-Ride Linear-Log

```{r}
lm(`Access-A-Ride` ~ log(All_Grades), weekly) %>% summary()
```

##### Bridges and Tunnels Linear-Log

```{r}
lm(Bridges_and_Tunnels ~ log(All_Grades), weekly) %>% summary()
```

\newpage

##### Subway and Buses Combined Linear-Log

```{r}
lm(MTA ~ log(All_Grades), weekly) %>% summary()
```

##### Bikes Linear-Log

```{r}
lm(Bikes ~ log(All_Grades), weekly) %>% summary()
```

\newpage

#### A.13

Linear-Log Regression Comparison

```{r}
weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ log(All_Grades), data = weekly)) %>%
  map(glance) %>%
  do.call(rbind.data.frame, .) %>%
  rownames_to_column() 



weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ log(All_Grades), data = weekly)) %>% 
  map(glance) %>% 
  do.call(rbind.data.frame, .) %>% 
  rownames_to_column %>% 
  as_tibble %>% 
  ggplot(aes(x = reorder(rowname, -adj.r.squared), y = adj.r.squared)) +
  geom_col() +
  coord_flip() + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Mode of Transportation",
       y = expression(paste("Adjusted ", R^{2})),
       title = expression(paste("Comparison of Adjusted ", R^{2})))
```

\newpage

#### A.14

Linear Regression with Dummy Variables

##### Subway Linear with log(Gas) + Trend + Season


```{r}
weekly_ts <- weekly %>%
  as_tsibble(index = Date)

model_subway <- weekly_ts %>% 
  model(TSLM(Subway ~ log(All_Grades) + trend() + fourier(K = 13)))

model_subway %>%
  report()
```

\newpage

##### Buses Linear with log(Gas) + Trend + Season

```{r}
model_buses <- weekly_ts %>%
  model(TSLM(Buses ~ log(All_Grades) + trend() + fourier(K = 13))) 

model_buses %>%
  report()
```

\newpage

##### LIRR Linear with log(Gas) + Trend + Season

```{r}
model_LIRR <- weekly_ts %>%
  model(TSLM(LIRR ~ log(All_Grades) + trend() + fourier(K = 13))) 

model_LIRR %>%
  report()
```

\newpage

##### MetroNorth Linear with log(Gas) + Trend + Season

```{r}
model_metronorth <- weekly_ts %>%
  model(TSLM(MetroNorth ~ log(All_Grades) + trend() + fourier(K = 13))) 

model_metronorth %>%
  report()
```

\newpage

##### Access-A-Ride Linear with Gas + Season

```{r}
weekly_ts %>% model(TSLM(`Access-A-Ride` ~ All_Grades + fourier(K = 13)))
```

\newpage

##### Bridges and Tunnels Linear with log(Gas) + Season

```{r}
weekly_ts %>% 
  model(TSLM(Bridges_and_Tunnels ~ log(All_Grades) + fourier(K = 13))) %>% 
  report()
```

\newpage

##### Combined Subway and Bus Linear with log(Gas) + Trend + Season

```{r}
weekly_ts %>% 
  model(TSLM(MTA ~ log(All_Grades) + trend() + fourier(K = 13))) %>%
  report()
```

\newpage

##### Bikes Linear with Gas + Season

```{r}
model_bikes <- weekly_ts %>%
  model(TSLM(Bikes ~ All_Grades + fourier(K = 13))) 

model_bikes %>%
  report()
```

\newpage

#### A.15 

Linear with Trend and Seasonality Regression Comparison

#### Linear with Trend and Seasonality

```{r}
model_aar <- weekly_ts %>% model(TSLM(`Access-A-Ride` ~ All_Grades + fourier(K = 13))) 

model_bat <- weekly_ts %>% model(TSLM(Bridges_and_Tunnels ~ log(All_Grades) + fourier(K = 13)))

model_mta <- weekly_ts %>% model(TSLM(MTA ~ log(All_Grades) + trend() + fourier(K = 13))) 

names <- c("Subway", "Buses", "LIRR", "MetroNorth", "Bikes", "Access-A-Ride",
           "Bridges_and_Tunnels", "MTA")

trend_models <- 
  rbind(glance(model_subway),
        glance(model_buses),
        glance(model_LIRR),
        glance(model_metronorth),
        glance(model_bikes),
        glance(model_aar),
        glance(model_bat),
        glance(model_mta)) %>%
  select(.model:adj_r_squared) %>%
  cbind(names, .)


trend_models %>% select(.model:adj_r_squared)


trend_models %>% 
  as_tibble %>% 
  ggplot(aes(x = reorder(names, -adj_r_squared), y = adj_r_squared)) +
  geom_col() +
  coord_flip() + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Mode of Transportation",
       y = expression(paste("Adjusted ", R^{2})),
       title = expression(paste("Comparison of Adjusted ", R^{2})))
```





