---
title: "The Effect of Gasoline Prices on Public Transit Ridership in New York City"
subtitle: "DATA 698"
author: 
  - Dominika Markowska-Desvallons
  - Orli Khaimova 
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
library(fpp3)
library(caret)
library(corrplot)
```

### Loading Data

INSERT DESCRIPTION OF DATA HERE

```{r}
mta2020 <- read.csv("https://raw.githubusercontent.com/okhaimova/DATA698/main/MTA_Daily_Ridership_Data__Beginning_2020.csv")
gas <- read.csv("https://raw.githubusercontent.com/okhaimova/DATA698/main/NYC%20Weekly%20Retail%20Gasoline%20and%20Diesel.csv",
                skip = 2, header = TRUE)
bicycle_count <- read.csv("Bicycle_Counts.csv")
counters <- read.csv("Bicycle_Counters.csv")
```

### Cleaning the Data

#### MTA Ridership Beginning 2020

DESCRIPTION

```{r}
mta2020 <- mta2020 %>%
  set_colnames(c("Date", "Subway", "Subway_Pre", "Buses", "Buses_Pre", "LIRR", "LIRR_Pre", 
                 "MetroNorth","MetroNorth_Pre", "Access-A-Ride", "Access-A-Ride_Pre", 
                 "Bridges_and_Tunnels", "Bridges_and_Tunnels_Pre")) %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y")) %>%
  arrange(Date) 
```



#### NYC Weekly Retail Gasoline and Diesel

DESCRIPTION

```{r}
gas <- gas %>%
  dplyr::select(1:9) %>%
  select(contains(c("Date","City"))) %>%
  set_colnames(c("Date", "All_Grades", "Regular", "Midgrade", "Premium")) %>%
  mutate(Date = as.Date(Date, format = "%b %d, %Y")) %>%
  na.omit()
```


#### Bicycle Count

DESCRIPTION
-date to get rid of times to sum by date
-explain filtering
-join from counter data
-comprehensive brooklyn bridge, comp manhattan, pulaski, ed koch, williamsburg
- The Brooklyn Bridge data seemed to have an outlier on February 8, 2022 with a reading of 
68,200, even though February 2022 had a daily average of 1348.2 bicycles per day. We chose to remove
the outlier and impute it with the daily average of February.

```{r, message = FALSE}
# these are the ids that match the five bridge counters
bridge_ids <- c("100010018", "100062893", "100009428", "300020904", "100009427")

counters <- counters %>%
  mutate(id = as.character(id))

bicycle_count <- bicycle_count %>%
  mutate(date = as.Date(date, "%m/%d/%Y")) %>%
  group_by(id, date) %>%
  summarise(total = sum(counts)) %>%
  filter(id %in% bridge_ids) %>%
  mutate(id = as.character(id)) %>%
  left_join(counters, by = 'id') %>%
  select(c(date, id, name, total)) %>%
  filter(date > "2014-12-31")

brooklyn_febraury_avg <- bicycle_count %>%
  filter(id == "300020904",
         date >= "2022-02-01" & date < "2022-03-01",
         date != "2022-02-08") %>%
  summarise(mean = mean(total)) %>%
  select(mean) %>%
  as.numeric()

# replace the outlier
bicycle_count <- bicycle_count %>%
  mutate(total = replace(total, total == 68200, round(brooklyn_febraury_avg,0)))
```

#### Combining Weekly

```{r, warning=FALSE, message=FALSE}
mta_weekly <- mta2020 %>%
  filter(Date > "2020-03-01") %>%
  mutate(Date = yearweek(Date)) %>%
  group_by(Date) %>%
  summarise(Subway = sum(Subway),
            Buses = sum(Buses),
            LIRR = sum(LIRR),
            MetroNorth = sum(MetroNorth),
            `Access-A-Ride` = sum(`Access-A-Ride`),
            Bridges_and_Tunnels = sum(Bridges_and_Tunnels)) %>%
  mutate(MTA = rowSums(across(c(Subway, Buses))))
  
gas_weekly <- gas %>%
  filter(Date > "2020-03-01") %>%
  mutate(Date = yearweek(Date))

bike_weekly <- bicycle_count %>%
  filter(date > "2020-03-01") %>%
  mutate(date = yearweek(date)) %>%
  group_by(date) %>%
  summarise(Bikes = sum(total)) %>%
  rename(Date = date) 


weekly <- plyr::join_all(list(mta_weekly, bike_weekly, gas_weekly), 
              by = "Date", type = 'left')

```


### Data Exploration

#### Gas

The price of gasoline was decreasing in the beginning of 2020, up until late February.
Gasoline prices started increasing as the World Health Organization (WHO) declared the novel 
Coronavirus (COVID-19) outbreak as a pandemic on March 11, 2020. 

It continued increasing, reaching the ultimate peak in June 2022. 

Some factors that affected the gasoline prices throughout the last few years were 
COVID-19 related supply disruptions, the Russian-Ukrainian conflict, and the Keystone XL
pipeline cancellation. 

The pattern in gasoline prices are similar across the different grades of gasoline.

```{r}
gas %>%
  filter(Date >= "2020-01-01") %>%
  ggplot(aes(x = Date, y = All_Grades)) +
  geom_line() +
  labs(title = "Price of Gasoline for All Grades, NYC",
       y = "$USD")

gas %>%
  filter(Date >= "2020-01-01") %>%
  pivot_longer(-Date) %>%
  filter(name != "All_Grades") %>%
  ggplot(aes(x = Date, y = value, colour = name)) +
  geom_line() +
  facet_grid(factor(name, levels =c("Regular", "Midgrade", "Premium")) ~., scales = "free_y") +
  labs(title = "Price of Gasoline, NYC",
       y = "$USD",
       color = "Gasoline Grade")
```


#### MTA

DESCRIPTION
-repeat weekly for other methods
-describe the charts
-repeat for bridges and tunnels
-fix scientific notation

```{r, message=FALSE}
mta2020 %>%
  dplyr::select(Date, Subway, Buses) %>%
  pivot_longer(-Date) %>%
  ggplot(aes(x = Date, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~., scales = "free_y") +
  labs(title = "MTA Ridership",
       y = "Count",
       color = "Mode")

mta2020 %>%
  dplyr::select(Date, LIRR, MetroNorth, "Access-A-Ride") %>%
  pivot_longer(-Date) %>%
  ggplot(aes(x = Date, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~., scales = "free_y") +
  labs(title = "MTA Ridership",
       y = "Count",
       color = "Mode")

mta2020 %>%
  mutate(Date = yearweek(Date)) %>%
  group_by(Date) %>%
  summarise(Subway = sum(Subway)) %>%
  as_tsibble(index = Date) %>%
  autoplot(Subway) +
  labs(title = "Weekly Subway Ridership",
       y = "Count")

mta2020 %>%
  as_tsibble(index = Date) %>%
  model(STL(Subway ~ season(window = "periodic"), robust = TRUE)) %>%
  components() %>%
  autoplot() +
  labs(title = "STL Decomposition")

  
```

#### Bicycle

-rename the labels

The five bridges seems to have a similar seasonality pattern. There is a peak of 
bicycle usage in the summer and a decline in the winter. Brooklyn Bridge, Manhattan Bridge,
and Williamsburg Bridge all connect Brooklyn and Manhattan. The Pulaski Bridge connects Brooklyn
and Queens. The Ed Koch Queensboro Bridge connects Queens and Manhattan. It is also the only bridge
that seems to have an apparent increase in usage from 2020 to the present.

```{r}
bicycle_count %>%
  ggplot(aes(x = date, y = total, color = name)) +
  geom_line() +
  facet_grid(name ~., scales = "free_y")

#dcmp <- bicycle_count %>%
#  as_tsibble(index = date, key = c(id, name)) %>%
#  fill_gaps() %>%
#  model(stl = STL(total))

weekly %>%
  as_tsibble(index = Date) %>%
  model(STL(Bikes ~ season(window = "periodic"), robust = TRUE)) %>%
  components() %>%
  autoplot() +
  labs(title = "STL Decomposition")
```


```{r}
weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = MTA), color = "steelblue") +
  geom_line(aes(y = All_Grades * 10000000), color = "darkred") +
  scale_y_continuous(name = "MTA Subway & Bus Ridership",
                     sec.axis = sec_axis(~./10000000, name = "Price of Gasoline")) +
  labs(title = "Price of Gasoline for All Grades, NYC")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = Buses), color = "steelblue") +
  geom_line(aes(y = All_Grades * 5000000), color = "darkred") +
  scale_y_continuous(name = "MTA Bus Ridership",
                     sec.axis = sec_axis(~./5000000, name = "Price of Gasoline")) +
  labs(title = "Price of Gasoline for All Grades, NYC")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = Subway), color = "steelblue") +
  geom_line(aes(y = All_Grades * 5000000), color = "darkred") +
  scale_y_continuous(name = "MTA Subway Ridership",
                     sec.axis = sec_axis(~./5000000, name = "Price of Gasoline")) +
  labs(title = "Price of Gasoline for All Grades, NYC")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = Bikes), color = "steelblue") +
  geom_line(aes(y = All_Grades * 100000), color = "darkred") +
  scale_y_continuous(name = "Bike Ridership",
                     sec.axis = sec_axis(~./100000, name = "Price of Gasoline")) +
  labs(title = "Price of Gasoline for All Grades, NYC")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = LIRR), color = "steelblue") +
  geom_line(aes(y = All_Grades * 500000), color = "darkred") +
  scale_y_continuous(name = "LIRR Ridership",
                     sec.axis = sec_axis(~./500000, name = "Price of Gasoline")) +
  labs(title = "Price of Gasoline for All Grades, NYC")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = MetroNorth), color = "steelblue") +
  geom_line(aes(y = All_Grades * 500000), color = "darkred") +
  scale_y_continuous(name = "MetroNorth Ridership",
                     sec.axis = sec_axis(~./500000, name = "Price of Gasoline")) +
  labs(title = "Price of Gasoline for All Grades, NYC")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = Bridges_and_Tunnels), color = "steelblue") +
  geom_line(aes(y = All_Grades * 1000000), color = "darkred") +
  scale_y_continuous(name = "Traffic",
                     sec.axis = sec_axis(~./1000000, name = "Price of Gasoline")) +
  labs(title = "Price of Gasoline for All Grades, NYC")

weekly %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = `Access-A-Ride`), color = "steelblue") +
  geom_line(aes(y = All_Grades * 50000), color = "darkred") +
  scale_y_continuous(name = "Traffic",
                     sec.axis = sec_axis(~./50000, name = "Price of Gasoline")) +
  labs(title = "Price of Gasoline for All Grades, NYC")
```

-back to work January 21
-omicron end of 21
```{r}
weekly %>%
  as_tsibble(index = Date) %>%
  model(classical_decomposition(Subway, type = "additive")) %>%
  components() %>%
  autoplot() +
  labs(title = "Classical Additive Decomposition of Subway Ridership")

lambda <- weekly %>%
  as_tsibble(index = Date) %>%
  features(Subway, features = guerrero) %>%
  pull(lambda_guerrero)

#stl decomp applied to the box cox transformed data
weekly %>%
  as_tsibble(index = Date) %>%
  model(STL(box_cox(Subway,lambda) ~ season(window = "periodic"), robust = TRUE)) %>%
  components() %>%
  autoplot() +
  ggtitle("STL with Box-Cox")

weekly %>%
  as_tsibble(index = Date) %>%
  gg_tsdisplay(difference(Subway), plot_type='partial') +
  labs(title = "hi")
  
```

### Modeling {.tabset}

#### Subway Linear

```{r}
model1<- lm(Subway ~ All_Grades, weekly)

summary(model1)

```

#### Subway Linear w/ Log Gas

```{r}
model2 <- lm(Subway ~ log(All_Grades), weekly)

summary(model2)


model2a<- lm(Subway ~ log(Regular) + log(Midgrade) + log(Premium), weekly)

summary(model2a)
```

#### MTA Linear w Log Gas 

```{r}
model3 <- lm(MTA ~ log(All_Grades), weekly)

summary(model3)


model3a<- lm(MTA ~ log(Regular) + log(Midgrade) + log(Premium), weekly)

summary(model3a)
```

#### Bus Linear

```{r}
model4<- lm(Buses ~ All_Grades, weekly)

summary(model4)
```

#### Log Bus Linear w/ Log Gas

```{r}
model5 <- lm(log(Buses) ~ log(All_Grades), weekly)

summary(model5)
```

#### LIRR Linear

```{r}
model6<- lm(LIRR ~ All_Grades, weekly)

summary(model6)

model6a<- lm(LIRR ~ Regular + Midgrade + Premium, weekly)

summary(model6a)

lm(LIRR ~ Regular + Midgrade + Premium, weekly) %>% summary()
```

#### LIRR Linear w/ Log Gas

```{r}
model7 <- lm(LIRR ~ log(All_Grades), weekly)

summary(model7)
```

#### MetroNorth Linear

```{r}
model8<- lm(MetroNorth ~ Regular + Midgrade + Premium, weekly)

summary(model8)
```

#### MetroNorth Linear w/ Log Gas

```{r}
model9 <- lm(MetroNorth ~ log(All_Grades), weekly)

summary(model9)
```

#### Bike Linear

```{r}
model10<- lm(Bikes ~ All_Grades, weekly)

summary(model10)

```

#### Bike Linear w/ Log Gas

```{r}
model11a<- lm(Bikes ~ log(Regular) + log(Midgrade) + log(Premium), weekly)

summary(model11a)
```



#### Subway Linear w/ Gas Trend Season

"An alternative to using seasonal dummy variables, especially for long seasonal periods, is to use Fourier terms.

If  m is the seasonal period

With Fourier terms, we often need fewer predictors than with dummy variables, especially when  
m is large. This makes them useful for weekly data, for example, where m is 52 "

```{r}
weekly_ts <- weekly %>%
  as_tsibble(index = Date)

model_subway <- weekly_ts %>% 
  model(TSLM(Subway ~ log(All_Grades) + trend() + fourier(K = 13)))

model_subway %>%
  report()
```

#### Buses Linear w/ Gas Trend Season


```{r}
model_buses <- weekly_ts %>%
  model(TSLM(Buses ~ log(All_Grades) + trend() + fourier(K = 13))) 

model_buses %>%
  report()
```

#### LIRR Linear w/ Gas Trend Season

```{r}
model_LIRR <- weekly_ts %>%
  model(TSLM(LIRR ~ log(All_Grades) + trend() + fourier(K = 13))) 

model_LIRR %>%
  report()
```

#### MetroNorth Linear w/ Gas Trend Season

```{r}
model_metronorth <- weekly_ts %>%
  model(TSLM(MetroNorth ~ log(All_Grades) + trend() + fourier(K = 13))) 

model_metronorth %>%
  report()
```

#### Bikes Linear w/ Gas Trend Season

-no trend in bikes, but big seasonality factor

```{r}
model_bikes <- weekly_ts %>%
  model(TSLM(Bikes ~ All_Grades + fourier(K = 13))) 

model_bikes %>%
  report()
```

### {-}

### Comparisons {.tabset}

#### Linear

```{r}
weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ All_Grades, data = weekly)) %>%
  map(glance) %>%
  do.call(rbind.data.frame, .) %>%
  rownames_to_column() 



weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ All_Grades, data = weekly)) %>% 
  map(glance) %>% 
  do.call(rbind.data.frame, .) %>% 
  rownames_to_column %>% 
  as_tibble %>% 
  ggplot(aes(x = reorder(rowname, -r.squared), y = r.squared)) +
  geom_col() +
  coord_flip() + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Mode of Transportation",
       y = expression( R^{2}),
       title = expression(paste("Comparison of ", R^{2})))
```

#### Linear Breakdown by Gas Type

```{r}
weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ Regular + Midgrade + Premium, data = weekly)) %>%
  map(glance) %>%
  do.call(rbind.data.frame, .) %>%
  rownames_to_column() 



weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ Regular + Midgrade + Premium, data = weekly)) %>% 
  map(glance) %>% 
  do.call(rbind.data.frame, .) %>% 
  rownames_to_column %>% 
  as_tibble %>% 
  ggplot(aes(x = reorder(rowname, -adj.r.squared), y = adj.r.squared)) +
  geom_col() +
  coord_flip() + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Mode of Transportation",
       y = expression(paste("Adjusted ", R^{2})),
       title = expression(paste("Comparison of Adjusted ", R^{2})))
```

#### Linear-Log 

```{r}
weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ log(All_Grades), data = weekly)) %>%
  map(glance) %>%
  do.call(rbind.data.frame, .) %>%
  rownames_to_column() 



weekly %>%
  select(-c(All_Grades, Regular, Midgrade, Premium, Date)) %>%
  map(~lm(.x ~ log(All_Grades), data = weekly)) %>% 
  map(glance) %>% 
  do.call(rbind.data.frame, .) %>% 
  rownames_to_column %>% 
  as_tibble %>% 
  ggplot(aes(x = reorder(rowname, -adj.r.squared), y = adj.r.squared)) +
  geom_col() +
  coord_flip() + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Mode of Transportation",
       y = expression(paste("Adjusted ", R^{2})),
       title = expression(paste("Comparison of Adjusted ", R^{2})))
```

#### Linear with Trend and Seasonality

```{r}
model_aar <- weekly_ts %>% model(TSLM(`Access-A-Ride` ~ All_Grades + fourier(K = 13))) 

model_bat <- weekly_ts %>% model(TSLM(Bridges_and_Tunnels ~ log(All_Grades) + fourier(K = 13)))

model_mta <- weekly_ts %>% model(TSLM(MTA ~ log(All_Grades) + trend() + fourier(K = 13))) 

names <- c("Subway", "Buses", "LIRR", "MetroNorth", "Bikes", "Access-A-Ride",
           "Bridges_and_Tunnels", "MTA")

trend_models <- 
  rbind(glance(model_subway),
        glance(model_buses),
        glance(model_LIRR),
        glance(model_metronorth),
        glance(model_bikes),
        glance(model_aar),
        glance(model_bat),
        glance(model_mta)) %>%
  select(.model:adj_r_squared) %>%
  cbind(names, .)


trend_models %>% select(.model:adj_r_squared)


trend_models %>% 
  as_tibble %>% 
  ggplot(aes(x = reorder(names, -adj_r_squared), y = adj_r_squared)) +
  geom_col() +
  coord_flip() + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Mode of Transportation",
       y = expression(paste("Adjusted ", R^{2})),
       title = expression(paste("Comparison of Adjusted ", R^{2})))
```



